---
title: "GTECH7852_23S_Week12_ac12980"
author: "Amy Carrillo"
date: "2023-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

libraries
```{r}
library(sf)
library(tidyverse)
library(mapview)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(leaflet)
library(leaflet.providers)
library(RColorBrewer)

```

```{r}
# Set working directory
wd <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(wd)
```


```{r}
acsPopByZip <- st_read("Data/R-Spatial_III_Lab/acsPopByZip.gpkg")
HW11 <- st_read("Data/R-Spatial_III_Lab/HW11.gpkg")
healthfacilities <- st_read("Data/R-Spatial_III_Lab/healthfacilities.gpkg")
```
## Task 1: 
### Plot at least two high-quality static maps with one using the COVID-19 data and one using a related factor. You can use either plot method for sf or ggplot method.

```{r}
#sf method
mypal <- brewer.pal(n=4, name= 'OrRd')

plot(acsPopByZip["Positiv"], 
     main = "Positive COVID 19 Cases", 
     col= mypal,
     border = NA,
     graticule = st_crs(4326),
     axes=TRUE,
     reset=FALSE)

```


```{r}
mypal <- brewer.pal(n=4, name= 'GnBu')
#ggplot method
plot(HW11["TotalPop"], 
     main = "NYC Total Population", 
     col=mypal,
     border = NA,
     graticule = st_crs(4326),
     axes=TRUE,
     reset=FALSE)
```

## Task 2: 
### Use ggplot2 and other ggplot-compatible packages to create a multi-map figure illustrating the possible relationship between COVID-19 confirmed cases or rate and another factor (e.g., the number of nursing homes, number of food stores, neighborhood racial composition, elderly population, etc.). The maps should be put side by side on one single page. Add graticule to at least one of those maps and label some of the feature on the map where applicable and appropriate.

### GGPLOT METHOD: Multimap 1
#### Positive COVID 19 Cases (4/12/2020 vs. 4/19/2020)

```{r}

# Creating breaks
br <- c(0, 25, 50, 75, 100) 

HW11$X4_12_20_PP<- cut(HW11$X4_12_20_cum_percpositive,
                              breaks = br,
                              dig.lab = 3)
HW11$X4_19_20_PP<- cut(HW11$X4_19_20_cum_percpositive,
                              breaks = br,
                              dig.lab = 3)
HW11$X4_23_21_PP<- cut(HW11$X4_23_21_cum_percpositive,
                              breaks = br,
                              dig.lab = 3)
```


```{r}

# Creating a palette
pal <- hcl.colors(6, "Inferno", rev = TRUE, alpha = 0.7)


plot1 <- ggplot() +
  geom_sf(data = HW11, 
          aes(fill= X4_12_20_PP)) +
          labs(
               title='Percent Positive 4/12/2020') +
          theme(legend.position = "none") +
          # Custom palette
          scale_fill_manual(values = pal,
                            drop = FALSE,
                            na.value = "grey80") 

plot2 <-  ggplot() + geom_sf(data = HW11, 
          aes(fill= X4_19_20_PP)) +
          labs( 
               title='Percent Positive 4/19/2020')+
          theme(legend.position = "none") +
          # Custom palette
          scale_fill_manual(values = pal,
                            drop = FALSE,
                            na.value = "grey80")+
          geom_sf_label(data=HW11 %>%
                filter(X4_12_20_cum_percpositive > 75),
                aes(label = X4_12_20_cum_percpositive),
                label.size = .01,
                size = 1.5)

grid.arrange(
  plot1,
  plot2,
  ncol = 2) 

```
### GGPLOT METHOD: Multimap 2
#### COVID-19 Cases and Health Facilties

```{r}


#clean
HW11$Total_HF <- HW11$Total_HealthFacilities
#palette
pal <- hcl.colors(6, "Inferno", rev = TRUE, alpha = 0.7)


p1 <-  ggplot() +
  geom_sf(data = acsPopByZip,
          aes(fill= Positiv)) +
          labs(
               title='Positive COVID-19 Cases')+
          #theme(legend.position = "none") +
          scale_fill_viridis_c(direction = -1)
          # Custom palette
          #scale_fill_manual(values = pal,
                            #drop = FALSE,
                            #na.value = "grey80") 
p2 <-  ggplot() + 
  geom_sf(data = HW11, 
          aes(fill= Total_HF)) +
          labs(
               title='Health Facilties') +
          #theme(legend.title = HF) + 
          # Custom palette
          scale_fill_viridis_c(direction = -1) 


grid.arrange(
  p1,
  p2,
  ncol = 2) 

```

## Task 3: 
### Create a web-based interactive map for COIVD-19 data using tmap, mapview, or leaflet package and save it as a HTML file.

### Interactive Web Map (Leaflet)
#### COVID 19 Positive Cases & Health Facilties

```{r}
# Creating a palete
pal_fun <- colorQuantile("OrRd", NULL, n = 5)

# Quick data clean-up
healthfacilities <- st_transform(healthfacilities, crs=4326) %>%
  filter(facility_latitude > 0)
acsPopByZip <- st_transform(acsPopByZip, crs=4326)

# Leaflet
polyHighlightOption <- leaflet::highlightOptions(opacity = 1.0, fillColor = 'black')
polyLabelOption <- leaflet::labelOptions(opacity = 0.6)
p_tip <- paste0("Positive Cases: ", acsPopByZip$Positiv);

htmlMap <- leaflet(acsPopByZip) %>%
  addPolygons(
    stroke = FALSE, 
    fillColor = ~pal_fun(Positiv),
    fillOpacity = 0.8, smoothFactor = 0.5,
    highlightOptions = polyHighlightOption,
    label = p_tip,
    labelOptions = polyLabelOption) %>%
  addCircles(radius = 0.02, weight = 1,
             data = healthfacilities %>% 
               sf::st_set_crs(4326) %>% sf::st_cast("POINT"),
             group = "Health Fac.") %>%
  addTiles(group = "OSM") %>%
  addProviderTiles("CartoDB.DarkMatter", group = "Carto") %>%
  addLayersControl(baseGroups = c("Carto", "OSM"), 
                   overlayGroups = c("COVID19", "HF")) %>%
  addLegend("bottomright",  # location
          pal=pal_fun,    # palette function
          values=~Positiv,  # value to be passed to palette function
          title = 'COVID-19 Positive Cases <br> & Health Facilities') # legend title


htmlMap

```

