---
title: "GTECH78520_23S_week11_24351622"
author: "Amy Carrillo"
date: "5-01-2023"
output: html_document
---

libraries
```{r}
library(sf)
library(tidyverse)
library(janitor)
library(stringr)
library(mapview)
library(purrr)

```

```{r}
# Set working directory
wd <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(wd)
```

### Part 1. Join the COVID-19 data to the NYC zip code area data (sf or sp polygons).

#### Adding data and data clean-up
```{r}
# Add NYC zip code shapefile
zipcode_data <- st_read("data/hw_data/ZIP_CODE_040114.shp")
zipcode_data <- clean_names(zipcode_data)
zipcode_data$zipcode = as.numeric(as.character(zipcode_data$zipcode))

# NYC zip code transform to 4326
zipcode_data %>%
  st_transform(4326)
```

```{r}
# Add COVID-19 test data csv files
COVID_TESTS_4_12_2020 <- read.csv("Data/R-Spatial_II_Lab/tests-by-zcta_2020_04_12.csv")
COVID_TESTS_4_19_2020 <- read.csv("Data/R-Spatial_II_Lab/tests-by-zcta_2020_04_19.csv")
COVID_TESTS_4_23_2021 <- read.csv("Data/R-Spatial_II_Lab/tests-by-zcta_2021_04_23.csv")

```

```{r}
# Clean names
COVID_TESTS_4_12_2020 <- clean_names(COVID_TESTS_4_12_2020)
COVID_TESTS_4_19_2020 <- clean_names(COVID_TESTS_4_19_2020)
COVID_TESTS_4_23_2021 <- clean_names(COVID_TESTS_4_23_2021)
```

```{r}
# Edit names to include the dates
colnames(COVID_TESTS_4_12_2020) <- c('zipcode','4_12_20_Positive','4_12_20_Total', '4_12_20_cum_percpositive')

colnames(COVID_TESTS_4_19_2020) <- c('zipcode','4_19_20_Positive','4_19_20_Total', '4_19_20_cum_percpositive')

```

```{r}
# Clean up 4_23_2021 data because it has more columns of data than the other covid datasets
COVID_TESTS_4_23_2021_select <- COVID_TESTS_4_23_2021 %>%
  select(modified_zcta, covid_case_count, total_covid_tests, percent_positive)

colnames(COVID_TESTS_4_23_2021_select) <- c('zipcode','4_23_21_Positive','4_23_21_Total', '4_23_21_cum_percpositive')
```

#### Joining dataframes (zipcode_data, COVID_TESTS_4_12_2020, COVID_TESTS_4_19_2020, COVID_TESTS_4_23_2021_select)

```{r}
# Join multiple data.frames
list_df = list(zipcode_data,COVID_TESTS_4_12_2020, COVID_TESTS_4_19_2020, COVID_TESTS_4_23_2021_select)

Zipcode_CovidTests <- list_df %>% 
  reduce(full_join, by='zipcode')%>%
  drop_na()%>%
  st_transform(crs = 4326)
```

#### Visual of joined dataframe
```{r}
mapview(Zipcode_CovidTests, zcol = c('4_12_20_cum_percpositive','4_19_20_cum_percpositive','4_23_21_cum_percpositive'), legend = FALSE)
```

### Part 2. Aggregate the NYC food retails store data (points) to the zip code data, so that we know how many retail stores in each zip code area. Note that not all locations are for food retail. And we need to choose the specific types according to the data.

#### Adding data and data clean-up for food retails store data
```{r}
# Add NYC food retails store data (points)
food_retails_xy <- read.csv("data/hw_data/nys_retail_food_store_xy.csv", fileEncoding = "Latin1", check.names = F)
```


```{r}
# Clean data and select only NYC boroughs
food_retails_xy <- clean_names(food_retails_xy)
food_retails_NY <- food_retails_xy %>%
  filter(zip_code > 7000) %>%
  filter(i_county %in% c("Bronx", "Kings","Queens", "New York","Richmond")) %>%
  filter(!is.na(x)) %>%
  filter(!is.na(y))

```


```{r}
# Turning csv into sf. Process the location column using stringr
st_as_sf(food_retails_NY %>% tidyr::drop_na(x, y), coords = c('x', 'y')) -> food_retailsNY_SF

# Assign coordinate system
st_crs(food_retailsNY_SF) <- 4326
```


```{r}
# First Trim off the extra white space in establishment_type columns
food_retailNY_SF2 <- food_retailsNY_SF %>% 
  mutate(across(where(is.character), str_trim))

# Filter the data to only food retails
ET = c("JAC", "JABC", "JABCP", "JACDK", "JABCDP", "JACD","JACFS", "JABCHK", "JACHK", "JABCK", "JACK", "JACDHK", "JACH", "JACDE", "JABCH", "JABCDH", "JABCD", "JACE", "JACI", "JACO ", "JACDH", "JABCG", "JACV", "JABCOP", "JAK ", "JACL", "JACG", "JABCDK", "JACZ", "JACW", "JCA", "JACDKM", "JABCGP", "JACN", "JABCKO", "JABCW", "JACDIK", "JACS", "JABCO", "JACDG", "JACP", "JABCHO", "JACHOP", "JACHO", "JACEW", "JDAC", "JKDAC", "JACHKO", "JACEK")

  
food_retailNY_SF3 <- food_retailNY_SF2 %>%
  filter(establishment_type %in% ET)

```

#### Find how many retail stores in each zip code area
```{r}
# Aggregate the NYC food retails store data (points) to the zip code data
foodretail_zip <- Zipcode_CovidTests %>%
  mutate(tract_area = st_area(geometry)) %>%
  st_transform (4326) %>%
  st_join(food_retailNY_SF3)

COVID_FoodRetail_ZIP <- foodretail_zip %>%
  select(c('zipcode','county', 'st_fips', 'cty_fips', 'po_name', '4_12_20_Positive','4_12_20_Total', '4_12_20_cum_percpositive','4_19_20_Positive','4_19_20_Total', '4_19_20_cum_percpositive','4_23_21_Positive','4_23_21_Total', '4_23_21_cum_percpositive'))

COVID_FoodRetail_ZIP <- COVID_FoodRetail_ZIP %>%
  group_by(zipcode) %>%
  mutate(Total_FoodRetail = n()) %>%
  distinct(zipcode, .keep_all = TRUE)

```

#### Visual for number of food retail stores by zipcode
```{r}
mapview(COVID_FoodRetail_ZIP, zcol = 'Total_FoodRetail')
```


### Part 3. Aggregate the NYC health facilities (points) to the zip code data. Similarly, choose appropriate subtypes such as nursing homes from the facilities.

#### Add data and data clean-up for health facilties
```{r}
# Add NYC health facilties data (points)
HealthFacilties_data <- read.csv("data/hw_data/NYS_Health_Facility.csv")
```


```{r}
# Clean data
HealthFacilties_data <- clean_names(HealthFacilties_data)
str(HealthFacilties_data)

# Filter data to only show NYC 5 boroughs
HealthFacilties_data <- HealthFacilties_data %>%
  filter(facility_county %in% c("Bronx", "Kings","Queens", "New York","Richmond"))

```

```{r}
# Turning health facilties csv into sf. 

#Process the location column using stringr
leftPos <- stringr::str_locate(HealthFacilties_data$facility_location, "\\(")[,1]
rghtPos <- stringr::str_locate(HealthFacilties_data$facility_location, "\\)")[,1]

# Get the coordinates text
HealthFacilties_data$facility_location %>% stringr::str_sub(leftPos+1, rghtPos -1) -> HealthFacilties_data$coords
cmmaPos <- stringr::str_locate(HealthFacilties_data$coords, ", ")

#Get the numeric coordinates
HealthFacilties_data$Y <- stringr::str_sub(HealthFacilties_data$coords, 1, cmmaPos[,1]-1) %>% as.numeric()
HealthFacilties_data$X <- stringr::str_sub(HealthFacilties_data$coords, cmmaPos[,2]+1) %>% as.numeric()

# Take out the rows without coordinates and make a sf object
st_as_sf(HealthFacilties_data %>% tidyr::drop_na(X, Y), coords = c('X', 'Y')) -> HealthFacilties_SF

# Assign coordinate system
st_crs(HealthFacilties_SF) <- 4326

```

```{r}
# Filter appropriate subtypes such as nursing homes from the facilities. 

# HF includes hospitals, clinics, and medical centers.
HF <- c("Diagnostic and Treatment Center","Hospital Extension Clinic","Hospital","Diagnostic and Treatment Center Extension Clinic","Primary Care Hospital - Critical Access Hospital Extension Clinic", "Primary Care Hospital - Critical Access Hospital")

NYC_HealthFacilties_SF <- HealthFacilties_SF %>%
  filter(description %in% HF)

```

#### Find how many health facilties in each zip code area
```{r}
# Aggregate the NYC health facilities (points) to the zip code data
COVID_FR_HealthFacilties_zip <- COVID_FoodRetail_ZIP%>%
  filter(zipcode >100) %>%
  mutate(tract_area = st_area(geometry)) %>%
  st_transform (4326) %>%
  st_join(NYC_HealthFacilties_SF) %>%
  select(c('zipcode','county', 'st_fips', 'cty_fips', 'po_name', '4_12_20_Positive','4_12_20_Total', '4_12_20_cum_percpositive','4_19_20_Positive','4_19_20_Total', '4_19_20_cum_percpositive','4_23_21_Positive','4_23_21_Total', '4_23_21_cum_percpositive','Total_FoodRetail')) %>%
  group_by(zipcode) %>%
  mutate(Total_HealthFacilities = n()) %>%
  distinct(zipcode, .keep_all = TRUE)

```

#### Data visualization for number of health faciltiies by zip code
```{r}
mapview(COVID_FR_HealthFacilties_zip, zcol = 'Total_HealthFacilities')
```

### Part 4. Join the Census ACS population, race, and age data to the NYC Planning Census Tract Data.

#### Adding data and data clean-up for NYC Planning Census Tract
```{r}
#Read NYC Planning Census Tract
nyc_census <- st_read("Data/R-Spatial_II_Lab/2010 Census Tracts/geo_export_1dc7b645-647b-4806-b9a0-7b79660f120a.shp") 

```
```{r}
# NYC Planning Census Tract: Create a county FIPS & census tract to later use for joining with ACS census data
nyc_census <- nyc_census %>%
  mutate(cntyFIPS = case_when(
    boro_name == 'Bronx' ~ '005',
    boro_name == 'Brooklyn' ~ '047',
    boro_name == 'Manhattan' ~ '061',
    boro_name == 'Queens' ~ '081',
    boro_name == 'Staten Island' ~ '085'),
    tractFIPS = paste(cntyFIPS, ct2010, sep=''))

```

#### Adding data and data clean-up for Census ACS data (population, race, and age data).
```{r}
# Read the Census ACS data
ACS_census <- read.csv("Data/R-Spatial_II_Lab/ACSDP5Y2018.DP05_data_with_overlays_2020-04-22T132935.csv")
```

```{r}
# Cleaning the data and selecting out population, race, and age data
ACS_census2 <- ACS_census[-1,]

#select out what we need
categories = c("DP05_0001E","DP05_0024E","DP05_0002E","DP05_0003E","DP05_0037E","DP05_0038E","DP05_0039E","DP05_0067E","DP05_0052E","DP05_0071E","DP05_0087E")

ACS_categories <- ACS_census2 %>%
  select(c("GEO_ID", "NAME", categories)) %>%
  rename("TotalPop"="DP05_0001E",
         "ElderlyPop" = "DP05_0024E",
         "MalePop" = "DP05_0002E",
         "FemalePop" = "DP05_0003E",
         "WhitePop" = "DP05_0037E",
         "BlackPop" = "DP05_0038E",
         "AmericanIndian" = "DP05_0039E",
         "AsianPop" = "DP05_0067E",
         "NativeHawaiian" = "DP05_0052E",
         "HispanicPop" = "DP05_0071E",
         "CitizenVotingAgePop" = "DP05_0087E") %>%
  mutate(census_code = str_sub(GEO_ID, -9))
```

#### Joining the NYC Planning Census Tract and ACS Census Data
```{r}
# Attribute Join. Join the Census ACS population, race, and age data to the NYC Planning Census Tract Data.
census_merge <- merge(nyc_census, ACS_categories, by.x = "tractFIPS", by.y = "census_code")

census_merge <- na.omit(census_merge)
```

#### Data visualization for NYC Planning Census Tract and ACS Census Data
```{r}
mapview(census_merge, zcol = "boro_name")
```


### Part 5: Aggregate the ACS census data to zip code area data.

#### Data set-up and clean-up
```{r}
# Create census centroids from census_merge file
census_centroids <- census_merge %>%
  st_centroid() %>%
  st_transform(4326) 
```
```{r}
# Make sure the dataframes are using  the same crs projection
zipcode_4326 <- COVID_FR_HealthFacilties_zip %>%
  st_transform(crs = 4326)

census_c4326 <- census_centroids %>%
  st_transform(crs = 4326)
```

```{r}
# Make sure the data is numeric
i <- c(16:25)
census_c4326[,i] <- apply(census_c4326[,i],2,
                      function(x) as.numeric(as.character(x)))
  
```

#### Join all of the census_merge data to zip code area data with the addition of demographics data
```{r}
census_centroid_zip_merge <- st_join(zipcode_4326, census_c4326, join = st_contains) %>%
  filter(zipcode > 100) %>%
  group_by(zipcode)
```

#### Consolidated version of the covid data, food retail stores, location, and census demographics information by zipcode
```{r}
nyc_zipcode_demongraphics <- st_join(zipcode_4326, census_c4326, join = st_contains) %>%
  filter(zipcode > 100) %>%
  select(c('zipcode','county', 'st_fips', 'cty_fips', 'po_name', '4_12_20_Positive','4_12_20_Total', '4_12_20_cum_percpositive','4_19_20_Positive','4_19_20_Total', '4_19_20_cum_percpositive','4_23_21_Positive','4_23_21_Total', '4_23_21_cum_percpositive','Total_FoodRetail','Total_HealthFacilities', 'TotalPop','ElderlyPop','MalePop','FemalePop','WhitePop', 'BlackPop','AmericanIndian','AsianPop','NativeHawaiian','HispanicPop')) %>%
  group_by(zipcode) %>%
  mutate(
        TotalPop = sum(TotalPop),
        ElderlyPop = sum(ElderlyPop),
        MalePop = sum(MalePop),
        FemalePop = sum(FemalePop),
        WhitePop = sum(WhitePop),
        BlackPop = sum(BlackPop),
        AmericanIndian = sum(AmericanIndian),
        AsianPop = sum(AsianPop),
        NativeHawaiian = sum(NativeHawaiian),
        HispanicPop = sum(HispanicPop)) %>%
  mutate(
        ElderlyPop_pp = ((ElderlyPop/TotalPop)*100),
        MalePop_pp = ((MalePop/TotalPop)*100),
        FemalePop_pp = ((FemalePop/TotalPop)*100),
        WhitePop_pp = ((WhitePop/TotalPop)*100),
        BlackPop_pp = ((BlackPop/TotalPop)*100),
        AmericanIndian_pp = ((AmericanIndian/TotalPop)*100),
        AsianPop_pp = ((AsianPop/TotalPop)*100),
        NativeHawaiian_pp = ((NativeHawaiian/TotalPop)*100),
        HispanicPop_pp = ((HispanicPop/TotalPop)*100)) %>%
  distinct(zipcode, .keep_all = TRUE)

```

```{r}
# clean datasets to get rid of NAs
census_centroid_zip_merge <- na.omit(census_centroid_zip_merge)
nyc_zipcode_demongraphics <-na.omit(nyc_zipcode_demongraphics)
```


#### Data visualization
```{r}
# Looking at the data
head(nyc_zipcode_demongraphics)
```

```{r}
# Testing out the new dataframe by looking at Hispanic population percentage rates by zipcode in NYC
mapview(nyc_zipcode_demongraphics, zcol = "HispanicPop_pp")
```

### Output
In the end, we should have the confirmed and tested cases of covid-19, numbers of specific types of food stores, numbers of specific types of health facilities, and population (total population, elderly, by race, etc.) at the zip code level. We should also have boroughs, names, etc. for each zip code area.

```{r}
# Confirmed and tested cases of covid-19
view(Zipcode_CovidTests)
# Numbers of specific types of food stores
view(COVID_FoodRetail_ZIP)
# Numbers of specific types of health facilities
view(COVID_FR_HealthFacilties_zip)
# MASTER DATASET. Population (total pop, race, elderly, etc.) at the zipcode level...We should also have boroughs, names, etc. for each zip code area.
view(nyc_zipcode_demongraphics)
```


